---
title: "Dply"
author: "Mats Hansson"
date: "2 april 2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown

```{r cars}

# load packages
```{r}
suppressMessages(library(dplyr))
library(nycflights13)
``````


## choosing columns: select rename

```{r}
# beside just using select() to pick columns...
flights %>%
  select (carrier, flight)

# you can use the minus sign to hide columns 
flights %>%
  select (-month, -day)
```


```{r}
# hide a range of columns
flights %>%
  select(-(year:carrier))

# hide any columns with a matching name
flights %>%
  select( -contains("time"))

```

```{r}
# pick a columns using a characther vector of columns name 
x<-c("carrier", "dest", "year")
flights %>%
  select(one_of(x))

```


```{r}
flights %>%
  rename(delay=arr_delay)

```


## choosing rows: filter, beteween, slice, sample_n, top_n, distinct


```{r}
# filter () support the use of multiple conditions 
flights %>%
  filter(dep_time<=520, dep_time>=500)

```

```{r}
# between() is a concise alternative for determining if numeric value fall in a range
flights %>%
  filter(between(dep_time, 500, 520))


# side note: is.na() can also be useful when filtering
flights %>%
  filter(!is.na(dep_time))

```


```{r}
# slice() filter rows by position
flights %>%
  slice(100: 150)


# keep the first three rows wihin each group 
flights %>% 
  group_by(month, day) %>%
  slice(1:3)


# sample three rows from each group
flights %>% group_by(month, day) %>% sample_n(3)

# keep three rows from each group with the top dep_delay
flights %>% group_by(month, day) %>% top_n(3, dep_delay)

# also sort by dep_delay within each group
flights %>% group_by(month, day) %>% top_n(3, dep_delay) %>% arrange(desc(dep_delay))

# unique rows can be identified using unique() from base R
flights %>% select(origin, dest) %>% unique()

# dplyr provides an alternative that is more "efficient"
flights %>% select(origin, dest) %>% distinct()

# side note: when chaining, you don't have to include the parentheses if there are no arguments
flights %>% select(origin, dest) %>% distinct


```


# Adding new variables: mutate, transmute, add_rownames

```{r}
# mutate() creates a new variable (and keeps all existing variables)
flights %>% mutate(speed = distance/air_time*60)

# transmute() only keeps the new variables
flights %>% transmute(speed = distance/air_time*60)
# example data frame with row names
mtcars %>% head()

# add_rownames() turns row names into an explicit variable
mtcars %>% add_rownames("model") %>% head()

# side note: dplyr no longer prints row names (ever) for local data frames
mtcars %>% tbl_df()

```


# Grouping and counting: summarise, tally, count, group_size, n_groups, ungroup

```{r}

# summarise() can be used to count the number of rows in each group
flights %>% group_by(month) %>% summarise(cnt = n())

# tally() and count() can do this more concisely
flights %>% group_by(month) %>% tally()
flights %>% count(month)

# you can sort by the count
flights %>% group_by(month) %>% summarise(cnt = n()) %>% arrange(desc(cnt))

# tally() and count() have a sort parameter for this purpose
flights %>% group_by(month) %>% tally(sort=TRUE)
flights %>% count(month, sort=TRUE)

# you can sum over a specific variable instead of simply counting rows
flights %>% group_by(month) %>% summarise(dist = sum(distance))

# tally() and count() have a wt parameter for this purpose
flights %>% group_by(month) %>% tally(wt = distance)
flights %>% count(month, wt = distance)

# group_size() returns the counts as a vector
flights %>% group_by(month) %>% group_size()

# n_groups() simply reports the number of groups
flights %>% group_by(month) %>% n_groups()

# group by two variables, summarise, arrange (output is possibly confusing)
flights %>% group_by(month, day) %>% summarise(cnt = n()) %>% arrange(desc(cnt)) %>% print(n = 10)

# ungroup() before arranging to arrange across all groups
flights %>% group_by(month, day) %>% summarise(cnt = n()) %>% ungroup() %>% arrange(desc(cnt))

```

# Creating data frames: data_frame

data_frame() is a better way than data.frame() for creating data frames. Benefits of data_frame():

You can use previously defined columns to compute new columns.
* It never coerces column types.
* It never munges column names.
* It never adds row names.
* It only recycles length 1 input.
* It returns a local data frame (a tbl_df).

```{r}
# data_frame() example
data_frame(a = 1:6, b = a*2, c = 'string', 'd+e' = 1) %>% glimpse()

# data.frame() example
data.frame(a = 1:6, c = 'string', 'd+e' = 1) %>% glimpse()
```

# Joining (merging) tables: left_join, right_join, inner_join, full_join, semi_join, anti_join

```{r}
# create two simple data frames
(a <- data_frame(color = c("green","yellow","red"), num = 1:3))
(b <- data_frame(color = c("green","yellow","pink"), size = c("S","M","L")))

# only include observations found in both "a" and "b" (automatically joins on variables that appear in both tables)
inner_join(a, b)

# include observations found in either "a" or "b"
full_join(a, b)

# include all observations found in "a"
left_join(a, b)

# include all observations found in "b"
right_join(a, b)

# right_join(a, b) is identical to left_join(b, a) except for column ordering
left_join(b, a)

# filter "a" to only show observations that match "b"
semi_join(a, b)

# filter "a" to only show observations that don't match "b"
anti_join(a, b)
# sometimes matching variables don't have identical names
b <- b %>% rename(col = color)

# specify that the join should occur by matching "color" in "a" with "col" in "b"
inner_join(a, b, by=c("color" = "col"))

```

# Viewing more output: print, View

```{r}
# specify that you want to see more rows
flights %>% print(n = 15)

# specify that you want to see ALL rows (don't run this!)
## flights %>% print(n = Inf)

# specify that you want to see all columns
flights %>% print(width = Inf)

# show up to 1000 rows and all columns
flights %>% View()

# set option to see all columns and fewer rows
options(dplyr.width = Inf, dplyr.print_min = 6)

# reset options (or just close R)
options(dplyr.width = NULL, dplyr.print_min = 10)

```

