---
title: " Going deeper with DPLYR"
author: "Mats Hansson"
date: "26 mars 2016"
output: html_document
---

## Dply function 

Great for data exploration  and transformation Inutatuve to write and easy to read, especially when using the "changing", syntax (covered below)




## dplyr functionality
Five basic verbs: "filter", "select", "arrange", "mutate", "summarise" (plus group_by)
Can work with data stored in databases and data tables
Joins: inner join, left join, semi-join, anti-join (not covered below)
Window functions for calculating ranking, offsets, and more
Better than plyr if you're only working with data frames (though it doesn't yet duplicate all of the plyr functionality)
Examples below are based upon the latest release, version 0.2 (released May 2014)

## Loading dplyr and an example dataset
dplyr will mask a few base functions
If you also use plyr, load plyr first
hflights is flights departing from two Houston airports in 2011

# load packages
```{r}
suppressMessages(library(dplyr))
library(hflights)


# explore data
data(hflights)
head(hflights)

```

* `tbl_sf` create a "local data frame" 
* LOcal data frame is simply a wrapper for a data frame that prints nicely



```{r}
# convert to a local data frame
flights<-tbl_df(hflights)
# printing only shows 10 rows and as many columns as can fit on your screen 
flights

```

```{r}
# you can specify that you want to see more rows 
print(flights, n=10)
```


```{r}
# Make a condition
filter(flights, Month==1, DayOfWeek==7)
```



```{r}
# To select more than one condition
filter(flights, UniqueCarrier %in% c("AA", "UA"))
filter(flights, TailNum %in% c("N576AA", "N504AA" ))
```


# Select columns by name 
```{r}
select(flights, Year, Month)

# use colons to select  multiple contigous columns and use `constains` to match columns by name
# note: `starts_with`, `ends_with and `matches` can also be used to match columns 

select(flights, Year, Month, contains("taxi"), contains("delay"))
select(flights, Year, ends_with("taxi"), starts_with("Taxi"))
select(flights, Year, matches("taxi"))
```


# Changing or piplining
* Usual way to perform multiple operations in one line is by nesting
* Can write commands in a natural order by using the `%>%` infix operator (which can be pronounced as "then")

```{r}
# chaining method
flights %>% select(UniqueCarrier, Year) %>% filter(Year==2011)
flights %>% select( Year, contains("Taxi")) %>% filter(Year==2011)
flights %>% select(DepTime) %>% filter(DepTime<1400)
flights %>% select (ArrTime) %>% filter(ArrTime %in% c(1500, 1501))
```

## arrange: Reorders rows 

```{r}
flights %>% 
  select(Year, Month) %>%
  arrange(desc(Month))

flights %>% select (ArrTime, TailNum) %>% arrange(ArrTime)
flights %>% select (ArrTime, TailNum) %>% arrange(desc(ArrTime))
flights %>% select(Year, contains("Taxi")) %>% arrange(desc(TaxiIn))
flights %>% select (Year,contains("Taxi")) %>% arrange(TaxiIn)
flights %>% select (Year,contains("Taxi")) %>% filter(TaxiIn %in% ("2")) %>% arrange(TaxiOut)


```

# Mutate: Add new variables 
* create a new variable that are functions of existing variables 

```{r}
flights %>% select(Distance, AirTime) %>% mutate(speed=Distance/AirTime*60)
flights<- flights %>% mutate(speed=Distance/AirTime*60)

test=flights %>%
  select(Distance, AirTime) %>%
  mutate(speed=Distance/AirTime*60)

```

# summarise: Reduce variables to values

* Primarily useful with data that has been grouped by one or more variables
* `group_by` creates the groups that will be operated on
* `summarise` uses the provided aggregation function to summarise each group

```{r}
# create a table grouped by Dest, and then summerise each group by taking the mean of ArrDelay.
flights %>%
  group_by(Dest) %>%
  summarise(avg_delay = mean(ArrDelay, na.rm=TRUE))


flights %>%
  group_by(DayOfWeek) %>%
  summarise(mean_AirTime=mean(AirTime, na.rm=TRUE))

```


* summarise_each allows you to apply the same summary function to multiple columns at once
* Note: mutate_each is also available

```{r}
# for each carrier, calculate the percentage of flights cancelled or diverted
flights %>% 
  group_by(Dest) %>%
  summarise(mean_cancelled=mean(Cancelled), mean_Diverted=mean(Diverted))

flights %>%
  group_by(Dest) %>%
  summarise_each(funs(mean), Cancelled, Diverted)


flights %>%
  group_by(Dest) %>%
  summarise(mean_Cancelled=mean(Cancelled,), na.rm=TRUE)


# for each carrier, calculate the minimum and maximum arrivala and departure delays
flights %>%
  group_by(UniqueCarrier) %>%
  summarise_each(funs(min(., na.rm=TRUE), max(.,na.rm=TRUE)), matches("Delay"))

flights %>% 
  group_by (Dest) %>%
  summarise_each(funs(mean(., na.rm=TRUE), sd(., na.rm=TRUE)), Cancelled, Diverted)
  
flights %>%
  group_by(TailNum) %>%
  summarise_each(funs(mean(., na.rm=TRUE), var(., na.rm=TRUE)), ArrTime, DepTime)

flights %>%
  group_by(Origin) %>%
  summarise_each(funs(mean(., na.rm=TRUE), sd(., na.rm=TRUE), min(., na.rm=TRUE), max(., na.rm=TRUE)), Distance, DepTime)

```

Helper function `n()` counts the number of rows in a group
Helper function `n_distinct(vector)` counts the number of unique items in that vector

```{r}
# for each day of the year, count the total number of flights and sort in descending order
flights %>%
  group_by(Month, DayofMonth) %>%
  summarise(flight_count=n()) %>%
  arrange(desc(flight_count))


flights %>%
  group_by(DayOfWeek, DayofMonth) %>%
  summarise(AirTime_n=n()) %>%
  arrange(desc(AirTime_n))

# rewrite more simply with the `tally` function 

flights %>%
  group_by(Month) %>%
  tally

flights %>%
  group_by(Month, DayofMonth) %>%
  tally()

# for each destination, count the total number of flights and the number of distinct planes that flew there
flights %>%
  group_by(Dest) %>%
  summarise(flight_count=n(), plane_count=n_distinct(TailNum))
```

* Grouping can sometimes be useful without summarsing 

```{r}
# For each destination, show the number of cancelled and not cancelled flights 

flights %>%
  group_by (Dest) %>%
  select(Cancelled) %>%
  table() %>%
  head()

flights %>%
  group_by(DayofMonth) %>%
  select(DayOfWeek) %>%
  table() %>%
  head()


```


## Window Functions
* Aggregation function (like `mean`) takes n inputs and returns 1 value
* [Window function] takes n inputs and returns n values
* Includes ranking and ordering functions (like `min_rank`), offset functions (`lead` and `lag`), and cumulative aggregates (like `cummean`).

```{r}
# for each carrier, calculate which two days of the year they had their longest departure delays
# note: smallest (not largest) value is ranked as 1, so you have to use `desc` to rank by largest value
flights %>%
  group_by(UniqueCarrier) %>%
  select(Month, DayofMonth, DepDelay) %>%
  filter(min_rank(desc(DepDelay))<=2) %>%
  arrange(UniqueCarrier, desc(DepDelay))

```


```{r}
# rewrite more simply with the `top_n` function
flights %>%
  group_by(UniqueCarrier) %>%
  select(Month, DayofMonth, DepDelay) %>%
  top_n(2) %>%
  arrange(UniqueCarrier, desc(DepDelay))


# for each month calculate the number if flights and the change from the previous month 
flights %>%
  group_by(Month) %>%
  summarise(flight_count=n()) %>%
  mutate(change=flight_count-lag(flight_count))

# rewrite more simply with the ´tally´ function
flights %>%
  group_by(Month) %>%
  tally() %>%
  mutate(change=n-lag(n))

```

## Other useful Convenience Functions 

```{r}
# randomly sample a fixed number of rows, without replacement 
flights %>% sample_n(5)

# randomly sample fraction of rows, with replacement 
flights %>% sample_n(5, replace=TRUE)

flights %>% sample_frac(0.5, replace=TRUE)


# dplyr approch: better formatting, and adapts to your screen width   
glimpse(flights)

```











